Analyzing portfolio effects in commercial catch data

### 2016-06-14 archive/catch_portfolios_mtg2.Rmd
More file rearranging

## Overview
We are interested in multiple metrics of diversity. For this first paper, we plan to divide the analysis into several components:  
1. Characterize trends in diversity over time (using GAMs). The metrics of diversity we might want to include are  
- variation in species diversity  
- variation in gear diversity  
- variation in spatial diversity (number of commercial areas fished?)  
- variation in permit holdings over time  
2. Using diversity as the predictor, look at absolute or relative change revenue / vessel (or revenue / vessel / days fished) as the response.

@@

```{r chunkSet, cache=FALSE, echo=FALSE,warning=FALSE,message=FALSE}
library(MASS)
library(knitr)
library(dplyr)
library(ggplot2)
library(vegan)
library(mgcv)
library(lme4)
library(scales)
library(fields)
library(reshape)
library(reshape2)
library(cluster)
opts_chunk$set(tidy=TRUE, message=FALSE)
```

## Data processing
Load the data from the saved workspace. This is the raw cfec data that has been merged to have the same columns across years, 1975 - present. The key columns are the species labels ("spec") and species number codes ("specn"), the gear number codes ("gearn"), the vessel id ("adfg"), the area where the catch occurred ("stat6"), the pounds harvested ("g_pounds"), gross earnings ("g_earn").

### Note: for any confidentiality concerns, the workspace loaded here has been slightly jittered. This will need to be replaced with the real data for the final model. 
```{r loadData}
load("cfec.Rdata")
# initially only use data 1985+
cfec = cfec[cfec$year>1984,]
```

The first data processing step is to aggregate some of the species. This is largely needed because of changes in reporting over time (before 1985 for example, lots of the catch was from foreign fleets & not reported to species level -- for example, some of the rockfishes). For other species, there are unidentified groups, so there's additional aggregation required.

```{r specLump}
# Lump all skates together
cfec$spec = as.character(cfec$spec)
cfec$specn[cfec$specn==700] = 700
cfec$specn[cfec$specn==701] = 700
cfec$specn[cfec$specn==702] = 700
cfec$spec[cfec$specn==700] = "SKATE"

# Lump all sharks together
cfec$specn[cfec$specn==691] = 689
cfec$specn[cfec$specn==689] = 689
cfec$spec[cfec$specn==689] = "SHARK"

# Lump flatfish (Shallow)
cfec$specn[cfec$specn==123] = 119 # rock
cfec$specn[cfec$specn==129] = 119 # starry flounder
cfec$specn[cfec$specn==127] = 119 # yellowfin
cfec$specn[cfec$specn==126] = 119 # butter
cfec$spec[cfec$specn==119] = "FLAT.shallow"
# 5, 6, 7, 15, 26, 47, 61, 
# Lump flatfish (deep)
cfec$specn[cfec$specn==124] = 118 # dover
cfec$specn[cfec$specn==134] = 118 # greenland turbot
cfec$spec[cfec$specn==118] = "FLAT.deep"

# Lump rockfish, codes from here:
cfec$specn[cfec$specn==141] = 144 # POP
cfec$specn[cfec$specn==136] = 144 # northern
cfec$specn[cfec$specn==151] = 144 # rougheye
cfec$specn[cfec$specn==166] = 144 # sharpchin
cfec$specn[cfec$specn==152] = 144 # shortraker
cfec$specn[cfec$specn==153] = 144 # redbanded
cfec$specn[cfec$specn==157] = 144 # silvergray
cfec$specn[cfec$specn==158] = 144 # redstripe
cfec$specn[cfec$specn==137] = 144 # bocaccio
cfec$specn[cfec$specn==144] = 144 # unid
cfec$spec[cfec$specn==144] = "ROCK.slope"

cfec$specn[cfec$specn==138] = 168 # copper
cfec$specn[cfec$specn==145] = 168 # yelloweye
cfec$specn[cfec$specn==146] = 168 # Canary
cfec$specn[cfec$specn==147] = 168 # quillback
cfec$specn[cfec$specn==148] = 168 # tiger
cfec$specn[cfec$specn==149] = 168 # China
cfec$specn[cfec$specn==150] = 168 # rosethorn
cfec$spec[cfec$specn==168] = "ROCK.demersal"

cfec$specn[cfec$specn==154] = 169 # dusky
cfec$specn[cfec$specn==155] = 169 # yellowtail
cfec$specn[cfec$specn==156] = 169 # widow
cfec$specn[cfec$specn==172] = 169 # dusky
cfec$specn[cfec$specn==173] = 169 # dark
cfec$specn[cfec$specn==142] = 169 # black
cfec$specn[cfec$specn==167] = 169 # blue
cfec$spec[cfec$specn==169] = "ROCK.pelagic"

# Lump king crabs
cfec$spec[cfec$spec%in%c("GKCR","BKCR","UKCR","RKCR")] = "KCRB" # king
# Lump tanner crabs
cfec$spec[cfec$spec%in%c("UTCR","BTCR")] = "TCRB" # tanner
# Lump shrimp - spot prawn, pink shrimp, 
cfec$spec[cfec$spec%in%c("SPRW","PSHP","USRM")] = "USRM" # shrimp
# Correct REX and "REX "
cfec$spec[cfec$spec%in%c("REX ")] = "REX"
# drop records with no spp
cfec = cfec[-which(cfec$spec==""),]
```

Next, we have to clean up a few of the gears. Drop a few gears that are sporadically used, and combine longline gears.

```{r}
# 8 = fishwheel, used at start / end
# 13 = dipnet, only last 2 years
# 99 = other, stopped reporting
cfec = cfec[-which(cfec$gearn%in%c(8, 13, 99)),]
# Merge 6 and 61 into longline
cfec$gearn[which(cfec$gearn==61)] = 6
```

Next, we'll deal with processing the areas. We have multiple types of area codes. The 6-digit codes represent groundfish management areas, the 5-digit codes are state salmon/shellfish areas, and the 3-digit codes are the halibut areas. For the groundfish and salmon areas, there are sub-areas that are also reported (as the last 2-digits), so we'll lump sub-areas together by stripping off the last 2-digits.

```{r}
# Filter out sub-districts and group things at larger areas
# Lump statistical areas by district -- groundfish
cfec$stat6 = as.numeric(cfec$stat6) # as.numeric needed because of integer
cfec$areaType = NA
cfec$stat6_char = NA
indx = which(cfec$stat6 > 99999 & cfec$stat6 < 999999)
cfec$stat6_char[indx] = nchar(as.character(cfec$stat6[indx]))
cfec$areaType[indx] = "groundfish"
# This is just turning the last digit - sub region - into 0
cfec$stat6[indx] = as.numeric(paste(substr(as.character(cfec$stat6[indx]),1,5), "0",sep=""))
# Now do salmon / shellfish - This is just turning the last 2 digits - sub region - into 00
indx = which(cfec$stat6 > 9999 & cfec$stat6 < 100000)
cfec$stat6[indx] = as.numeric(paste(substr(as.character(cfec$stat6[indx]), 1, 3), "00", sep=""))
cfec$areaType[indx] = "salmonShellfish"

# Deal with the IPHC areas too. We can verify that nearly all spp caught in 3-digit areas are halibut,
#sort(table(cfec$spec[which(cfec$stat6 < 1000)]))
```

Next, we'll aggregate the stat6 codes into larger geographic regions. The focus here is on Prince William Sound, but for contrast we'll include Southeast Alaska, Kodiak W / E, Cook Inlet, and Bristol Bay.

```{r}
# Add the areas -- 5 digit are salmon/shellfish, 6-digit are groundfish
cfec$area = NA
cfec$area[cfec$stat6==485900] = "PWS"
cfec$area[cfec$stat6==475900] = "PWS"
cfec$area[cfec$stat6==465900] = "PWS"
cfec$area[cfec$stat6==455900] = "PWS"
cfec$area[cfec$stat6==445900] = "PWS"
cfec$area[cfec$stat6==495930] = "PWS"
cfec$area[cfec$stat6==485930] = "PWS"
cfec$area[cfec$stat6==475930] = "PWS"
cfec$area[cfec$stat6==465930] = "PWS"
cfec$area[cfec$stat6==455930] = "PWS"
cfec$area[cfec$stat6==445930] = "PWS"
cfec$area[cfec$stat6==486000] = "PWS"
cfec$area[cfec$stat6==476000] = "PWS"
cfec$area[cfec$stat6==466000] = "PWS"
cfec$area[cfec$stat6==456000] = "PWS"
cfec$area[cfec$stat6==446000] = "PWS"
cfec$area[cfec$stat6==486030] = "PWS"
cfec$area[cfec$stat6==476030] = "PWS"
cfec$area[cfec$stat6==466030] = "PWS"
cfec$area[cfec$stat6==456030] = "PWS"
cfec$area[cfec$stat6==486100] = "PWS"
cfec$area[cfec$stat6==476100] = "PWS"
cfec$area[cfec$stat6==466100] = "PWS"
cfec$area[which(cfec$stat6 < 20099 & cfec$stat6 >= 20000)] = "PWS"
cfec$area[which(cfec$stat6 < 21299 & cfec$stat6 >= 21200)] = "PWS"
cfec$area[which(cfec$stat6 < 22799 & cfec$stat6 >= 22700)] = "PWS"  
cfec$area[which(cfec$stat6 < 22699 & cfec$stat6 >= 22600)] = "PWS"
cfec$area[which(cfec$stat6 < 22899 & cfec$stat6 >= 22800)] = "PWS"
cfec$area[which(cfec$stat6 < 22199 & cfec$stat6 >= 22100)] = "PWS"
cfec$area[which(cfec$stat6 < 22299 & cfec$stat6 >= 22200)] = "PWS"  
cfec$area[which(cfec$stat6 < 22599 & cfec$stat6 >= 22500)] = "PWS"
cfec$area[which(cfec$stat6 < 22499 & cfec$stat6 >= 22400)] = "PWS"
cfec$area[which(cfec$stat6 < 22399 & cfec$stat6 >= 22300)] = "PWS" 
# Add halibut areas
cfec$area[cfec$stat6 %in% c(242,232,220,230,240)] = "PWS"

cfec$area[cfec$stat6==525900] = "CookInlet"
cfec$area[cfec$stat6==535900] = "CookInlet"
cfec$area[cfec$stat6==515930] = "CookInlet"
cfec$area[cfec$stat6==525930] = "CookInlet"
cfec$area[cfec$stat6==535930] = "CookInlet"
cfec$area[cfec$stat6==516000] = "CookInlet"
cfec$area[cfec$stat6==526000] = "CookInlet"
cfec$area[cfec$stat6==496030] = "CookInlet"
cfec$area[cfec$stat6==506030] = "CookInlet"
cfec$area[cfec$stat6==516030] = "CookInlet"
cfec$area[cfec$stat6==526030] = "CookInlet"
cfec$area[cfec$stat6==506100] = "CookInlet"
cfec$area[cfec$stat6==516100] = "CookInlet"
cfec$area[which(cfec$stat6 < 24799 & cfec$stat6 >= 24400)] = "CookInlet" 
cfec$area[which(cfec$stat6%in%c(261, 272))] = "CookInlet"

cfec$area[which(cfec$stat6 < 11599 & cfec$stat6 >= 11100)] = "Southeast"
cfec$area[which(cfec$stat6 < 10499 & cfec$stat6 >= 10100)] = "Southeast"
cfec$area[which(cfec$stat6 < 11099 & cfec$stat6 >= 10500)] = "Southeast" 
cfec$area[which(cfec$stat6 >= 305430 & cfec$stat6 <= 305530)] = "Southeast"
cfec$area[which(cfec$stat6 >= 315430 & cfec$stat6 <= 315600)] = "Southeast"
cfec$area[which(cfec$stat6 >= 325430 & cfec$stat6 <= 325700)] = "Southeast"
cfec$area[which(cfec$stat6 >= 335430 & cfec$stat6 <= 335800)] = "Southeast"
cfec$area[which(cfec$stat6 >= 345400 & cfec$stat6 <= 345830)] = "Southeast"
cfec$area[which(cfec$stat6 >= 355400 & cfec$stat6 <= 355900)] = "Southeast"
cfec$area[which(cfec$stat6 > 365400 & cfec$stat6 < 365830)] = "Southeast"
# For IPHC see Fig. 20 here: http://www.iphc.int/publications/techrep/tech0049.pdf
cfec$area[which(cfec$stat6%in%c(140, 141, 142, 143, 144, 150, 151, 152, 153, 160, 161, 162, 163, 170, 171, 172, 174, 174, 181, 182, 183, 184, 185))] = "Southeast"

# Kodiak east area is defined by halibut areas 270/280, includes ADFG salmon / 
# shellfish on E side of kodiak, and groundfish areas extending to NMFS Area 630
# boundary
cfec$area[cfec$stat6%in%c(270,280)] = "Kodiak.east"
cfec$area[which(cfec$stat6 < 25899 & cfec$stat6 >= 25800)] = "Kodiak.east"
cfec$area[which(cfec$stat6 < 25999 & cfec$stat6 >= 25900)] = "Kodiak.east"
cfec$area[cfec$stat6==535630] = "Kodiak.east"
cfec$area[cfec$stat6==525630] = "Kodiak.east"
cfec$area[cfec$stat6==515630] = "Kodiak.east"
cfec$area[cfec$stat6==505630] = "Kodiak.east"
cfec$area[cfec$stat6==495630] = "Kodiak.east"
cfec$area[cfec$stat6==535700] = "Kodiak.east"
cfec$area[cfec$stat6==525700] = "Kodiak.east"
cfec$area[cfec$stat6==515700] = "Kodiak.east"
cfec$area[cfec$stat6==505700] = "Kodiak.east"
cfec$area[cfec$stat6==495700] = "Kodiak.east"
cfec$area[cfec$stat6==525730] = "Kodiak.east"
cfec$area[cfec$stat6==515730] = "Kodiak.east"
cfec$area[cfec$stat6==505730] = "Kodiak.east"
cfec$area[cfec$stat6==495730] = "Kodiak.east"
cfec$area[cfec$stat6==525800] = "Kodiak.east"
cfec$area[cfec$stat6==515800] = "Kodiak.east"
cfec$area[cfec$stat6==505800] = "Kodiak.east"
cfec$area[cfec$stat6==495800] = "Kodiak.east"

cfec$area[cfec$stat6%in%c(271,281)] = "Kodiak.west"
cfec$area[cfec$stat6==535730] = "Kodiak.west"
cfec$area[cfec$stat6==545730] = "Kodiak.west"
cfec$area[cfec$stat6==525800] = "Kodiak.west"
cfec$area[cfec$stat6==535800] = "Kodiak.west"
cfec$area[cfec$stat6==545800] = "Kodiak.west"
cfec$area[cfec$stat6==535830] = "Kodiak.west"
cfec$area[cfec$stat6==525830] = "Kodiak.west"
cfec$area[which(cfec$stat6 < 25699 & cfec$stat6 >= 25500)] = "Kodiak.west"
cfec$area[which(cfec$stat6 < 25499 & cfec$stat6 >= 25300)] = "Kodiak.west"  
cfec$area[which(cfec$stat6 < 26299 & cfec$stat6 >= 26000)] = "Kodiak.west"

# Bristol Bay, defined as going west to the boundary between NMFS areas 512 / 516
cfec$area[cfec$stat6%in%c(31600, 31700, 31800, 32000, 32100, 32200, 32400, 32500, 32600)] = "BristolBay"
cfec$area[cfec$stat6%in%c(575730, 575800, 575830)] = "BristolBay"
cfec$area[cfec$stat6%in%c(585700, 585730, 585800, 585830)] = "BristolBay"
cfec$area[cfec$stat6%in%c(595630, 595700, 595730, 595800, 595830)] = "BristolBay"
cfec$area[cfec$stat6%in%c(605600, 605630, 605700, 605730, 605800, 605830)] = "BristolBay"
cfec$area[cfec$stat6%in%c(615600, 615630, 615700, 615730, 615800, 615830)] = "BristolBay"

# Area to add - Alaska peninsula - Halibut areas 300-340?
```

@@

```{r}
T = table(cfec$adfg, cfec$stat6)
CC = clara(T, k = 8)
cluster_vesselArea = t(CC$medoids)
```

@@

```{r}
specArea = table(cfec$spec, cfec$area)
# normalize by area
for(i in 1:dim(specArea)[2]) {
specArea[,i] = specArea[,i]/sum(specArea[,i])
}
kable(round(specArea[-which(apply(specArea,1,max) < 0.01),], 2))
```

## Exploratory data analysis

Drop all species records that have less than 1000 total observations. This is discarding about 5000 of 20 million records. 
```{r discardSpec}
cfecClean = group_by(cfec, spec) %>%
  mutate(nspecies = n()) %>%
  filter(nspecies > 1000) %>%
  data.frame
# Also drop out unid groundfish, unid flatfish, smelts, grenadier, english sole -- all very rare
cfecClean = cfecClean[-which(cfecClean$spec%in%c("UFLT","UGRN","SMLT","GRDR","EGLS")),]
```

@@

# Write the SSL related sites to a workspace
sslStat6 = c(355630,
365730, 375830, 345530, 335430, 335430, 335430,
335430, 335430, 365800, 355700, 345530, 355630,
355700, 335700, 335700, 355700, 445930, 495930, 
476000, 485930, 495930, 325600, 495930, 476000,
475930, 555530, 565600, 525700, 525700, 525830,
525700, 515800, 515830, 535630, 365800, 575630,
535630, 565630, 525830, 595500, 595430, 595500,
595430, 635430, 635430, 605430, 615430, 635504,
585500)

# calculate areas in circle around square of interest
sslStat6.square = c(sslStat6, sslStat6 +- 9970, sslStat6 +- 10000, 
sslStat6 +- 10070, sslStat6 +- 70, sslStat6 +- 9930, sslStat6 +- 10000, 
sslStat6 +- 10030, sslStat6 +- 30)
sslStat6.label = rep(sslStat6,9)

cfecSSL = cfecClean[cfecClean$stat6%in%sslStat6.square,]
cfecSSL$label = sslStat6.label[match(cfecSSL$stat6, sslStat6.square)]
save(cfecSSL, file="cfecSSL.Rdata")

@@

```{r runningMean}
totalPWSExp$nspecies = (totalPWSExp$nspecies + (totalPWSExp$nspecies + totalPWSExp$diffspec))/2
totalPWSExp$ngear = (totalPWSExp$ngear + (totalPWSExp$ngear + totalPWSExp$diffgear))/2
totalPWSExp$narea = (totalPWSExp$narea + (totalPWSExp$narea + totalPWSExp$diffarea))/2
totalPWSExp$narea = (totalPWSExp$narea + (totalPWSExp$narea + totalPWSExp$diffarea))/2
```

Start with simple linear model, no random effects  
```{r linearModel}
lm.simple = lm(difflog~as.factor(year) + nspecies + ngear + narea + diffspec + diffgear + diffarea,data=totalPWSExp)
summary(lm.simple)
```

@@

```{r simpleGam}
gam.simple = gam(difflog~s(year) + nspecies + ngear + narea + diffspec + diffgear + diffarea,data=totalPWSExp)
summary(gam.simple)
```

@@

First, it's useful to look at the number of vessels that can be uniquely assigned to each of our areas. This shows that > 80% of vessels never fish in more than 1 of our coarse areas (Prince William Sound, Cook Inlet, Bristol Bay, E/W Kodiak, etc), so we can start our analysis focusing on those.  
```{r tableVesselArea}
tableVesselArea = table(cfecClean$adfg, cfecClean$area)
tableVesselArea = ceiling(tableVesselArea/1.0e10)
T = table(apply(tableVesselArea,1,sum))
m = matrix(NA, 3, 2)
colnames(m) = c("# Vessels", "Percent")
rownames(m) = c("1","2","3+")
m[,1] = c(T[2],T[3],sum(T[-c(1:3)]))
m[,2] = round(100*m[,1]/sum(m[,1]),3)

@@

Next, we'll extract the vessels that are only assigned to one area, subsetting the entire dataset to just these vessels.
```{r }
singleAreaVessels = rownames(tableVesselArea)[which( apply(tableVesselArea, 1, sum) == 1)]
cfecClean$adfg = as.character(cfecClean$adfg)
cfecSingles = group_by(cfecClean, adfg) %>% 
  mutate(is_Single = any(adfg %in% singleAreaVessels)) %>%
  as.data.frame() %>%
  filter(is_Single) %>%
  data.frame
```

We can re-run the code above to total the response / predictors by calendar year. We'll add a column for area. 
```{r aggregateSingleVessel}
totalSingles = group_by(cfecSingles, adfg, year) %>%
  summarize(totalearnings = sum(g_earn), 
            nspecies = length(unique(spec)),  
            ngear = length(unique(gearn)),  
            narea = length(unique(area)), 
            nfishery = length(unique(p_fshy))) 
# Lump larger groups
totalSingles$nspecies[totalSingles$nspecies > 10] = 10
totalSingles$narea[totalSingles$narea > 2] = 2
totalSingles$ngear[totalSingles$ngear > 3] = 3
totalSingles$nfishery[totalSingles$nfishery > 4] = 4

@@

```{r simpleAreaGam, fig.pos="placeHere",fig.cap="Simple GAM model, grouped by area, with area specific effects of species and change in species fished. Gears, change in gears, areas, and change in areas fished also included. Plots show mean smooth year term, with SE."}
gam.area = gam(difflog~area + s(year,by=as.factor(area)) + area*nspecies + area*diffspec + ngear + diffgear,data=totalSingles)
summary(gam.area)
par(mfrow = c(3,2),mgp=c(2,1,0),mai=c(0.5,0.5,0.3,0.1))
plot(gam.area)
```

@@

******
# More rigorous analysis

@@

```{r}
cfecCleaned = cfecClean
cfecCleaned$isNA = 0 # is an area ever NA?
cfecCleaned$isNA[is.na(cfecCleaned$area)] = 1
vessel.isna = table(cfecCleaned$adfg,cfecCleaned$isNA)
cfecCleaned = cfecCleaned[cfecCleaned$adfg %in% names(which(vessel.isna[, 2] ==
0)), ]
```

```{r}
# Of this restricted dataset, 93% of vessels don't move between areas -
# ever. So the primary focus of our analysis can be on gears and species
# landed. Just so the area switching won't matter, let's delete those
# records that have multiple areas
t = ceiling(table(cfecCleaned$adfg, cfecCleaned$area)/1e+10)
# delete vessels that fish in 2+ areas
cfecCleaned = cfecCleaned[cfecCleaned$adfg %in% names(which(apply(t, 1, sum) ==
1)), ]

@@

# Lump larger groups
totalVessels$nspecies[totalVessels$nspecies > 10] = 10
totalVessels$ngear[totalVessels$ngear > 3] = 3
totalVessels$nfishery[totalVessels$nfishery > 4] = 4
```

@@

```{r}
dummyDF = expand.grid(adfg = unique(totalVessels$adfg), 
  year = min(totalVessels$year) : max(totalVessels$year),
  stringsAsFactors = F)

>>>>

# Overview
We are interested in multiple metrics of diversity. For this first paper, we plan to divide the analysis into several components:  
1. Characterize trends in diversity over time (using GAMs). The metrics of diversity we might want to include are  
- variation in species diversity  
- variation in gear diversity  
- variation in spatial diversity (number of commercial areas fished?)  
- variation in permit holdings over time  
2. Using diversity as the predictor, look at absolute or relative change revenue / vessel (or revenue / vessel / days fished) as the response.

@@

```{r chunkSet, cache=FALSE, echo=FALSE,warning=FALSE,message=FALSE}
library(MASS)
library(knitr)
library(dplyr)
library(ggplot2)
library(vegan)
library(mgcv)
library(lme4)
library(scales)
library(fields)
library(reshape)
library(reshape2)
library(cluster)
opts_chunk$set(tidy=TRUE, message=FALSE)
load("cfecCleaned.Rdata")
```

@@

```{r}
if(useADFG == TRUE) T = table(cfec$adfg, cfec$stat6)
if(useADFG == FALSE) T = table(cfec$cfecid, cfec$stat6)
CC = clara(T, k = 8)
cluster_vesselArea = t(CC$medoids)
```

@@

```{r}
specArea = table(cfec$spec, cfec$area)
# normalize by area
for(i in 1:dim(specArea)[2]) {
specArea[,i] = specArea[,i]/sum(specArea[,i])
}
# look at only species that make up > 1% in any single area
kable(round(specArea[-which(apply(specArea,1,max) < 0.01),], 2))
```

## Identifying data problems with permits  
Doing some additional sorting, we can see that there are a few unique ADFG numbers that have a huge number of permits associated with them. These are likely data issues, and should be removed.  
```{r}
head(group_by(cfec,adfg) %>% 
       summarise(permits=length(unique(p_serial))) %>% 
       arrange(-permits) %>% 
       data.frame,20)
```

So we can delete these extreme permit holders. Per R. Brenner / J. Shriver, many of these are research vessels, out of state groundfish, PSC-related bycatch, etc.

```{r}
cfec = cfec[-which(cfec$adfg >= 98000),]
cfec = cfec[-which(cfec$adfg == 0),]
cfec = cfec[-which(is.na(cfec$adfg)),] # drop NAs
```

# Exploratory data analysis (Fall 2015)

Drop all species records that have less than 1000 total observations. This is discarding about 5000 of 20 million records. 
```{r discardSpec}
cfecClean = group_by(cfec, spec) %>%
  mutate(nspecies = n()) %>%
  filter(nspecies > 1000) %>%
  data.frame
# Also drop out unid groundfish, unid flatfish, smelts, grenadier, english sole -- all very rare
cfecClean = cfecClean[-which(cfecClean$spec%in%c("UFLT","UGRN","SMLT","GRDR","EGLS")),]
cfecClean = cfecClean[-which(cfecClean$gearn%in%c(0, 18, 22, 27, 37, 41, 62, 85)), ]
```

@@

# Write the SSL related sites to a workspace
#sslStat6 = c(355630,
#365730, 375830, 345530, 335430, 335430, 335430,
#335430, 335430, 365800, 355700, 345530, 355630,
#355700, 335700, 335700, 355700, 445930, 495930, 
#476000, 485930, 495930, 325600, 495930, 476000,
#475930, 555530, 565600, 525700, 525700, 525830,
#525700, 515800, 515830, 535630, 365800, 575630,
#535630, 565630, 525830, 595500, 595430, 595500,
#595430, 635430, 635430, 605430, 615430, 635504,
#585500)

# calculate areas in circle around square of interest
#sslStat6.square = c(sslStat6, sslStat6 +- 9970, sslStat6 +- 10000, 
#sslStat6 +- 10070, sslStat6 +- 70, sslStat6 +- 9930, sslStat6 +- 10000, 
#sslStat6 +- 10030, sslStat6 +- 30)
#sslStat6.label = rep(sslStat6,9)

#cfecSSL = cfecClean[cfecClean$stat6%in%sslStat6.square,]
#cfecSSL$label = sslStat6.label[match(cfecSSL$stat6, sslStat6.square)]
#save(cfecSSL, file="cfecSSL.Rdata")

# Just for general interest, aggregate earnings by year and vessel
# Similar to Kasperski and Holland
#earnings_byVesselYear = aggregate(cfecClean$g_earn, by = list("Vessel"=cfecClean$adfg, "Year"= cfecClean$year), FUN = sum, na.rm=T)
#median(earnings_byVesselYear$x) # 40884
#hist(log(earnings_byVesselYear$x), 40, col="grey")
```

@@

```{r runningMean,include=FALSE,echo=FALSE}
totalPWSExp$nspecies = (totalPWSExp$nspecies + (totalPWSExp$nspecies + totalPWSExp$diffspec))/2
totalPWSExp$ngear = (totalPWSExp$ngear + (totalPWSExp$ngear + totalPWSExp$diffgear))/2
totalPWSExp$narea = (totalPWSExp$narea + (totalPWSExp$narea + totalPWSExp$diffarea))/2
totalPWSExp$narea = (totalPWSExp$narea + (totalPWSExp$narea + totalPWSExp$diffarea))/2
```

Start with simple linear model, no random effects  
```{r linearModel,include=FALSE,echo=FALSE}
lm.simple = lm(difflog~as.factor(year) + nspecies + ngear + narea + diffspec + diffgear + diffarea,data=totalPWSExp)
summary(lm.simple)
```

@@

```{r simpleGam,include=FALSE,echo=FALSE}
gam.simple = gam(difflog~s(year) + nspecies + ngear + narea + diffspec + diffgear + diffarea,data=totalPWSExp)
summary(gam.simple)
```

@@

First, it's useful to look at the number of vessels that can be uniquely assigned to each of our areas. This shows that > 80% of vessels never fish in more than 1 of our coarse areas (Prince William Sound, Cook Inlet, Bristol Bay, E/W Kodiak, etc), so we can start our analysis focusing on those.  
```{r tableVesselArea,include=FALSE,echo=FALSE}
tableVesselArea = table(cfecClean$adfg, cfecClean$area)
tableVesselArea = ceiling(tableVesselArea/1.0e10)
T = table(apply(tableVesselArea,1,sum))
m = matrix(NA, 3, 2)
colnames(m) = c("# Vessels", "Percent")
rownames(m) = c("1","2","3+")
m[,1] = c(T[2],T[3],sum(T[-c(1:3)]))
m[,2] = round(100*m[,1]/sum(m[,1]),3)

@@

Next, we'll extract the vessels that are only assigned to one area, subsetting the entire dataset to just these vessels.
```{r ,include=FALSE,echo=FALSE}
singleAreaVessels = rownames(tableVesselArea)[which( apply(tableVesselArea, 1, sum) == 1)]
cfecClean$adfg = as.character(cfecClean$adfg)
cfecSingles = group_by(cfecClean, adfg) %>% 
  mutate(is_Single = any(adfg %in% singleAreaVessels)) %>%
  as.data.frame() %>%
  filter(is_Single) %>%
  data.frame
```

We can re-run the code above to total the response / predictors by calendar year. We'll add a column for area. 
```{r aggregateSingleVessel,include=FALSE,echo=FALSE}
totalSingles = group_by(cfecSingles, adfg, year) %>%
  summarize(totalearnings = sum(g_earn), 
            nspecies = length(unique(spec)),  
            ngear = length(unique(gearn)),  
            narea = length(unique(area)), 
            nfishery = length(unique(p_fshy))) 
# Lump larger groups
totalSingles$nspecies[totalSingles$nspecies > 10] = 10
totalSingles$narea[totalSingles$narea > 2] = 2
totalSingles$ngear[totalSingles$ngear > 3] = 3
totalSingles$nfishery[totalSingles$nfishery > 4] = 4

@@

```{r simpleAreaGam, fig.pos="placeHere",fig.cap="Simple GAM model, grouped by area, with area specific effects of species and change in species fished. Gears, change in gears, areas, and change in areas fished also included. Plots show mean smooth year term, with SE.",include=FALSE,echo=FALSE}
#gam.area = gam(difflog~area + s(year,by=as.factor(area)) + area*nspecies + area*diffspec + ngear + diffgear,data=totalSingles)
#summary(gam.area)
#par(mfrow = c(3,2),mgp=c(2,1,0),mai=c(0.5,0.5,0.3,0.1))
#plot(gam.area)
```

@@

******
\newpage

# More rigorous analysis (Spring 2016)

@@

```{r}
cfecCleaned = cfecClean
cfecCleaned$isNA = 0 # is an area ever NA?
cfecCleaned$isNA[is.na(cfecCleaned$area)] = 1
if(useADFG==TRUE) {
  vessel.isna = table(cfecCleaned$adfg,cfecCleaned$isNA)
  cfecCleaned = cfecCleaned[cfecCleaned$adfg %in% names(which(vessel.isna[, 2] ==
0)), ]  
}

if(useADFG==FALSE) {
  vessel.isna = table(cfecCleaned$cfecid,cfecCleaned$isNA)
  cfecCleaned = cfecCleaned[cfecCleaned$cfecid %in% names(which(vessel.isna[, 2] ==
0)), ] 
}

```{r}
if(useADFG == TRUE) {
# Of this restricted dataset, 93% of vessels don't move between areas -
# ever. So the primary focus of our analysis can be on gears and species
# landed. Just so the area switching won't matter, let's delete those
# records that have multiple areas
t = ceiling(table(cfecCleaned$adfg, cfecCleaned$area)/1e+10)
# delete vessels that fish in 2+ areas
cfecCleaned = cfecCleaned[cfecCleaned$adfg %in% names(which(apply(t, 1, sum) ==
1)), ]

# also make a table of continuous vessels
T = ceiling(table(cfecCleaned$adfg, cfecCleaned$year)/1.0e10)
contVessels = names(which(apply(T,1,sum)==dim(T)[2]))

@@

contVess.df = aggregate(cbind(nspecies,ngear,nfishery,totalearnings) ~ adfg, data = totalVessels[totalVessels$adfg%in%contVessels, ], FUN = mean)
contVess.df$cv = aggregate(totalearnings ~ adfg, data = totalVessels[totalVessels$adfg%in%contVessels, ], FUN = sd)$totalearnings / contVess.df$totalearnings

# Lump larger groups
totalVessels$nspecies[totalVessels$nspecies > 10] = 10
totalVessels$ngear[totalVessels$ngear > 3] = 3
totalVessels$nfishery[totalVessels$nfishery > 4] = 4
} # end useADFG

```{r}
if(useADFG == FALSE) {
# Of this restricted dataset, 93% of vessels don't move between areas -
# ever. So the primary focus of our analysis can be on gears and species
# landed. Just so the area switching won't matter, let's delete those
# records that have multiple areas
t = ceiling(table(cfecCleaned$cfecid, cfecCleaned$area)/1e+10)
# delete vessels that fish in 2+ areas
cfecCleaned = cfecCleaned[cfecCleaned$cfecid %in% names(which(apply(t, 1, sum) ==
1)), ]

# also make a table of continuous vessels
T = ceiling(table(cfecCleaned$cfecid, cfecCleaned$year)/1.0e10)
contVessels = names(which(apply(T,1,sum)==dim(T)[2]))

# Now tabulate summaries of species, gears, fisheries, etc
totalVessels = group_by(cfecCleaned, cfecid, year) %>%  
  summarize(totalearnings = sum(g_earn),  
  nspecies = length(unique(spec)),  
  ngear = length(unique(gearn)),  
  nfishery = length(unique(p_fshy)),
  area = area[1])

# calculate landings by species, year, vessel
landBySpec = group_by(cfecCleaned, cfecid, year, spec) %>%  
summarize(totalpounds = sum(g_pounds))  
# calculate effective number of species by exp() for shannon
# http://jonlefcheck.net/2012/10/23/diversity-as-effective-numbers/
landBySpec = group_by(landBySpec, cfecid, year) %>%  
   summarize(effectiveSpec = exp(diversity(totalpounds)))

# calculate landings by gear, year, vessel
landByGear = group_by(cfecCleaned, cfecid, year, gearn) %>%  
  summarize(totalpounds = sum(g_pounds))  
landByGear = group_by(landByGear, cfecid, year) %>%  
  summarize(effectiveGear = exp(diversity(totalpounds)))

# calculate landings by fishery, year, vessel
landByFshy = group_by(cfecCleaned, cfecid, year, p_fshy) %>%  
summarize(totalpounds = sum(g_pounds))  
landByFshy = group_by(landByFshy, cfecid, year) %>%  
summarize(effectiveFshy = exp(diversity(totalpounds)))

contVess.df = aggregate(cbind(nspecies,ngear,nfishery,totalearnings) ~ cfecid, data = totalVessels[totalVessels$cfecid%in%contVessels, ], FUN = mean)
contVess.df$cv = aggregate(totalearnings ~ cfecid, data = totalVessels[totalVessels$cfecid%in%contVessels, ], FUN = sd)$totalearnings / contVess.df$totalearnings

# Lump larger groups
totalVessels$nspecies[totalVessels$nspecies > 10] = 10
totalVessels$ngear[totalVessels$ngear > 3] = 3
totalVessels$nfishery[totalVessels$nfishery > 4] = 4
} # end useADFG

@@

```{r}
if(useADFG==TRUE) {
dummyDF = expand.grid(adfg = unique(totalVessels$adfg), 
  year = min(totalVessels$year) : max(totalVessels$year),
  stringsAsFactors = F)
